---
title: "Logistic Regression R Notebook (2)"
author: "Rachel M. Smith"
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache=FALSE}
source("SETUP.R")
options(width = 70)
knitr::opts_chunk$set(
    tidy = FALSE,
    echo = TRUE,
    cache = FALSE,
    fig.keep = 'high',
    fig.show = 'asis',
    results = 'asis',
    # autodep = T,
    Rplot = TRUE,
    dev = 'pdf',
    fig.path = 'graphics/reuters/rplot_',
    fig.width = 7,
    fig.asp = 1,
    out.width = "\\linewidth",
    collapse = TRUE
)
```

# Data Summary

```{r dat}
dat <- read.spss("data/reuters.sav", to.data.frame = TRUE)
dat$party <- sapply(dat$party, R.na, v = 99)
dat$response <- recode_factor(dat$response, 
                              "other/no opinion" = NA_character_)
dat <- na.omit(dat)
summary(dat)

```

# Logistic Regression Model

$$ \ln\left(\frac{\pi}{1 - \pi}\right) = \alpha + \beta X $$

$$ \pi = \frac{e^{\alpha + \beta X}}{1 + e^{\alpha + \beta X}} $$


```{r lrm}
lrm <- glm(response ~ ind, data = dat, family = "binomial")
# summary(lrm)
xtable::xtable(lrm)
nulldev <- lrm$null.deviance
nulldev.df <- lrm$df.null
nulldev.p <- c(round(nulldev, 2), nulldev.df)
# **Null Deviance**
dev <- lrm$deviance
dev.df <- lrm$df.residual
dev.p <- c(round(dev, 2), dev.df)
# **Observed Deviance**
aic <- lrm$aic
aic.p <- c(round(aic, 2), " ")
# **AIC**
fits <- as.data.frame(rbind(nulldev.p, dev.p, aic.p))
rownames(fits) <- c("**Null Deviance**", "**Residual Deviance**", "**AIC**")
names(fits) <- c("Estimate", "Degrees of Freedom")
```

```{r echo=FALSE}
kable(fits, caption = "Logistic Regression Model Fit Statistics")

R.msmm(residuals(lrm))[-5] %>%
    kable(caption = "Logistic Regression: Deviance Residuals Summary [note]") %>%
    add_footnote("_**M** = Mean, **SD** = Standard Deviation, **Min** = Minimum, \\& **Max** = Maximimum_", threeparttable = TRUE)
```

## Logit Model: Confidence Intervals (CI) \& Odds Ratios (OR)

`r tufte::newthought("Odds Ratio")`


```{r lrmOR}
OR <- exp(coef(lrm)) 
```


`r tufte::newthought("Confidence Intervals.")` The _first CI computed below (`'CI.b'`)_ is that of the _standardized coefficient ($\beta$)_, which is based on the coefficient's _standard errors (SE)_. The _second computed CI (`'CI.phi'`_ is based on the logistic regression model's _profiled log-likelihood function_, making it the _odds ratio ($\Phi$) CI_. 


```{r lrmCI}
## CIs using standard errors ##
CI.b <- confint(lrm, trace = FALSE)

## CIs using profiled log-likelihood ##
CI.phi <- confint(lrm)

```



```{r echo=FALSE}
CIb <- cbind(coef(lrm), CI.b)
CIb.labs <- c("**$\\beta$**", dimnames(confint(lrm))[[2]])
CIb <- rbind(CIb.labs, round(CIb, 4))
dimnames(CIb)[[1]][1] <- " "
dimnames(CIb)[[2]] <- c(" ", "$CI_{\\beta}$", " ")

## odds ratios and 95% CI ##
ORCI <- exp(cbind(coef(lrm), CI.phi))
ORCI.labs <- c("**$\\Phi$**", dimnames(confint(lrm))[[2]])
ORCI <- rbind(ORCI.labs, round(ORCI, 4))
dimnames(ORCI)[[1]][1] <- " "
dimnames(ORCI)[[2]] <- c(" ", "$CI_{\\Phi}$", " ")


kable(CIb, align = rep('r', ncol(CIb)), 
      caption = "Logistic Regression Coefficients ($\\beta$) \\& 
      Coresponding Confidence Intervals (_CI_)")

kable(ORCI, align = rep('r', ncol(ORCI)), 
      caption = "Logistic Regression Odds Ratios ($\\Phi$) \\& 
      Coresponding Confidence Intervals (_CI_) [note]") %>%
    add_footnote("Confidence intervals are based on the logistic regression 
                 model's profiled log-likelihood function, 
                 rather than the standard errors", 
                 threeparttable = TRUE)
```


## Wald's Chi-Square Test


```{r lrmchsq1}
library(aod) ## "wald.test()" ##
wald.test(b = coef(lrm), Sigma = vcov(lrm), Terms = 2)
```


\newpage
# References`r R.cite_r("~/GitHub/auxDocs/REFs.bib", footnote = TRUE)`

