---
title: "HW-2: Question-3 R Notebook"
author: "Rachel M. Smith"
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache=FALSE}
source("../SETUP.R")
knitr::opts_chunk$set(
    tidy = TRUE,
    echo = TRUE,
    cache = FALSE,
    fig.keep = 'high',
    fig.show = 'asis',
    results = 'asis',
    autodep = T,
    Rplot = TRUE,
    dev = 'pdf',
    fig.path = 'graphics/HW2/rplot_',
    fig.width = 7,
    fig.height = 7,
    out.width = "\\linewidth"
    )
```

# Question-3 Instructions


The data for the remaing sets of analyses are from a national telephone interview study of batterer intervention program (BIP) standards advisory and compliance monitoring committees. Respondents were asked a series of questions varying in structure from _open-ended_ to simple _yes-or-no_. These analyses will concern the discrete data collected in response to the following interview questions:

> How many members currently serve on your standards committee?
> Does your organization have methods for assessing programs’ feedback about the standards?
> Do the standards apply to programs designed for all genders?

The primary interest for the below analyses relate to an overarching effort to implement and sustain effective and appropriate anti-violence intervention strategies among female-identified perpetrators of same-sex violence. The above listed questions provide a mix of continuous numeric and dichotomous (`0 = No`; `1 = Yes`) indicators of responding states' current organizational and ideological capacities for such intervention strategies.

-----

```{r q3Dat}
dat <- read.csv("data/states_new.csv")
# datmap <- read.csv("data/usmap.csv")

dat <- dat[, c("Q3", "Q7", "Q8")]
dat <- na.omit(dat)
```

```{r results='asis', fig.margin=TRUE}
dat <- subset(dat, Q7 < 2)
dat <- subset(dat, Q8 < 2)
dat.f <- within(dat, {## could also do this with an "apply()" function ##
    Assessments <- factor(Q7, levels = c(0, 1), labels = c("No", "Yes"))
    Gender.Inclusive <- factor(Q8, levels = c(0, 1), labels = c("No", "Yes"))
})

# dat.c <- within(dat, {## could also do this with an "apply()" function ##
#     Q7 <- Q7 - mean(Q7)
#     Q3 <- Q3 - mean(Q3)
# })

library(mediation) ## "mediations()" & "plot.mediations()" ##

set.seed(42) ## i'll explain later ##
treatment <- c("Q3")
mediators <- c("Q7")
outcome <- c("Q8")
datasets <- list(dat = dat)
m.model <- mediations(datasets, treatment, mediators, outcome,
                      families = c("gaussian", "binomial"),
                      interaction = TRUE,
                      conf.level = .95, sims = 50)

mod.y <- m.model$Q3.Q7.Q8[["model.y"]]
mod.m <- m.model$Q3.Q7.Q8[["model.m"]]

mod.y
mod.m

plot(m.model, labels = c("Indirect", "Direct", "Total"))
```


```{r intEffects, echo=TRUE, results='asis'}
## "interaction.plot()" from the {stats} package ##
with(dat.f, {
    interaction.plot(
    Gender.Inclusive,
    Assessments,
    Q7,
    # col = c("darkgray", "#cd0000"),
    lwd = 2,
    main = "Interaction Effects",
    # ylab = expression(mu[welfare]),
    # xlab = "Abuse",
    # trace.label = "Program"
    )
})
```
