---
title: "HW-2: Question-1 R Notebook"
author: "Rachel M. Smith"
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache=FALSE}
source("../SETUP.R")
knitr::opts_chunk$set(
    tidy = FALSE,
    echo = FALSE,
    cache = FALSE,
    fig.keep = 'high',
    fig.show = 'hold',
    results = 'hide',
    autodep = T,
    Rplot = TRUE,
    dev = 'pdf',
    fig.path = 'graphics/HW2/rplot-q1-',
    fig.width = 7,
    fig.height = 7,
    out.width = "\\linewidth"
    )
```

-----

`r tufte::newthought("Data Summary.")` [Portland Police 2014 report on racial profiling in traffic stops](http://www.portlandoregon.gov/police/article/481668).

```{r dat}
dat <- R.rspss("data/race2.sav") 
## My "R.rspss()" function is a convenience wrapper function for 
    ## "read.spss()" {pkg:foreign} with the default values for 
    ## "use.value.labels" & "to.data.frame" set to "FALSE" and "TRUE", 
    ## respectively. ## 

dat <- R.rspss("data/race2.sav") 
## My "R.rspss()" function is a convenience wrapper function for 
    ## "read.spss()" {pkg:foreign} with the default values for 
    ## "use.value.labels" & "to.data.frame" set to "FALSE" and "TRUE", 
    ## respectively. ## 

dat <- within(dat, {## I like to keep the a numeric and factor copy 
                        ## of variables. Below, I'm defining the factor
                        ## variables for "ped" and "offense" according to 
                        ## their original labels in the spss datafile. ##
    offense <- sapply(offense, R.na, v = 0)
    ped.f <- factor(ped, 
                    levels = unique(ped), 
                    labels = c("African American", 
                               "Asian", 
                               "Hispanic", 
                               "Native American", 
                               "White"))
    offense.f <- factor(offense, 
                        levels = c(1, 2), 
                        labels = c("major", 
                                   "minor"))
})

s.dat <- R.msmm(dat)
## My "R.msmm()" function is another convenience function I wrote, which 
    ## calculates the mean, standard deviation, minimum, and maximum
    ## (hence "msmm") values and a count of existing NAs for each 
    ## numeric or integer column in a given dataframe and returns a summative 
    ## dataframe of these values. ##

s.dat

ped.t <- table(dat$ped.f) %>% as.data.frame() 
## dataframe of factor level counts for the "ped.f" factor variable ##
offense.t <- table(dat$offense.f) %>% as.data.frame()
## dataframe of factor level counts for the "offense.f" factor variable ##

kable(ped.t, col.names = c("Pedestrian Race", "Frequency"))
kable(offense.t, col.names = c("Offense", "Frequency"))

dat.ft <- ftable(dat$ped, dat$offense) %>% 
    as.matrix()
## The "ftable()" function {pkg:base} produces a flat contingency table, and 
    ## is thus highly convenient for extracting joint frequencies ##
dimnames(dat.ft) <- list("Pedestrian Race" = paste(levels(dat$ped.f)), 
                         "Offense" = c("NA", 
                                     paste(levels(dat$offense.f)))
                         )
dat.ft
    ## Shows that the NAs (i.e., '0's) in dat$offense are for 
        ## Asian, Hispanic, and Native American; but not 
        ## White nor African American, cases. ##
    ## This means that we can just remove all cases (rows) with NAs, 
        ## since we are only interested in the White and African American 
        ## cases for this analysis. ##

dat <- na.omit(dat) %>% 
    droplevels() 
## ^- remove cases with "NAs" throughout the dataframe, then 
    ## drop the any resultingly unused levels of the factor variables ##
```


```{r mA}
m.A <- matrix(c("$n_{11}$", "$n_{12}$", "**$n_{1+}$**",
                "$n_{21}$", "$n_{22}$", "**$n_{2+}$**",
                "**$n_{+1}$**", "**$n_{+2}$**", "_**$n_{++}$**_"), nrow = 3, byrow = TRUE)
dimnames(m.A) <- list("_**I**_" = c("$i_{1}$", "$i_{2}$", "$i+$"),
                      "_**J**_" = c("$j_{1}$", "$j_{2}$", "$+j$"))
noquote(m.A)
```
  
```{r ft}
# str(ft.m)
ft <- ftable(dat$ped.f, dat$offense.f) %>% 
    as.matrix()
dimnames(ft) <- list(ped = dimnames(ft)[[1]], 
                     offense = dimnames(ft)[[2]])
n <- sum(ft)
pi.j <- apply(ft, 1, sum) %>% matrix()
rownames(pi.j) <- rownames(ft)
pi.i <- apply(ft, 2, sum) %>% matrix(nrow = 1, ncol = 2)
colnames(pi.i) <- colnames(ft)
pi.i <- cbind(pi.i, n)

ft.j <- cbind(ft, pi.j)

ft.m <- rbind(ft.j, pi.i)

pT <- ft.m[[9]]

AA <- ft.m[[7]]
pAA <- paste0(round((AA/pT)*100, digits = 2), "%")
WH <- ft.m[[8]]
pWH <- paste0(round((WH/pT)*100, digits = 2), "%")

MA <- ft.m[[4]]
pMA <- paste0(round((MA/pT)*100, digits = 2), "%")
MI <- ft.m[[6]]
pMI <- paste0(round((MI/pT)*100, digits = 2), "%")

AAMA <- ft.m[[1]]
pAAMA <- paste0(round((AAMA/pT)*100, digits = 2), "%")
WHMA <- ft.m[[2]]
pWHMA <- paste0(round((WHMA/pT)*100, digits = 2), "%")
AAMI <- ft.m[[3]]
pAAMI <- paste0(round((AAMI/pT)*100, digits = 2), "%")
WHMI <- ft.m[[5]]
pWHMI <- paste0(round((WHMI/pT)*100, digits = 2), "%")

```

```{r ftmmllm, results='asis'}
ft.mm <- ft.m
dimnames(ft.mm) <- list(Race = c("_African American_", 
                                 "_White_", 
                                 "$\\pi_{i+}$"), 
                       Offense = c("_Major_", 
                                   "_Minor_", 
                                   "$\\pi_{+j}$"))
ft.mm <- matrix(ft.mm, nrow = 3, byrow = T, 
                dimnames = list(Race = c("**African American**", 
                                         "**White**", 
                                         "**$\\pi_{i+}$**"), 
                                Offense = c("Major", 
                                            "Minor", 
                                            "$\\pi_{+j}$")))

kable(ft.mm, caption = "Contingency Table of Pedestrians Stopped by Portland Police: Pedestrian Race (2 Levels) x Severity of Offense (2 Levels).")
```


```{r tbl, fig.margin=TRUE, results='asis'}
tbl <- with(dat, {
    table(ped.f, offense.f)
})
dimnames(tbl) <- list(ped.f = levels(dat$ped.f),
                      offense.f = levels(dat$offense.f))
par(family = "serif")
mosaicplot(tbl, type = "pearson", color = pal_my.a75[c(5, 16)], 
           main = "Cross-Tabulation of Pedestrian Race X Offense",
           xlab = "Pedestrian Race", ylab = "Offense Severity", cex = 1)
```

# $1.a.$ Loglinear Model^[\complete]

$$ \log(\mu_{ij}) = \lambda + \lambda_{i}^{X} + \lambda_{j}^{Y} + \lambda_{ij}^{XY} $$

-----

`r tufte::newthought("Loglinear Analysis Summary.")` A loglinear model was used to test the association between pedestrian race and whether individuals were stopped by Portland Police (_PPD_) for major versus minor offenses. `r pWH` of pedestrians stopped by PPD were White, and `r pAA` were African American ($N_{Total} = `r pT`$). Among the `r AA` African American pedestrians, `r pAAMA` (`r AAMA`) were stopped for major offenses, while `r pAAMI` (`r AAMI`) were stopped for minor offenses. Among the `r WH` White pedestrians, `r pWHMA` (`r WHMA`) were stopped for major offenses, while `r pWHMI` (`r WHMI`) were stopped for minor offenses. Results from the likelihood ratio test of the $2 x 2$ contingency table (see above) indicated no significant differences regarding these factors, $G^{2}(1) = 0.14, p = 0.70$.

-----


```{r llm, results='asis'}
library(MASS) ## "loglm()" (... and many other useful functions) ##
llm <- loglm( ~ ped.f + offense.f, data = tbl)
llm
```


\newpage



# $1.b.$ Logistic Regression Model^[\complete]

$$ \ln\left(\frac{\pi}{1 - \pi}\right) = \alpha + \beta X $$

$$ \pi = \frac{e^{\alpha + \beta X}}{1 + e^{\alpha + \beta X}} $$


```{r lrm}
lrm <- glm(offense.f ~ ped, data = dat, family = "binomial", x = TRUE)
summary(lrm)
lrm.b <- coef(lrm)
lrm.b1 <- coef(lrm)[[2]] ## \beta_{1} (i.e., "ped") ##
lrm.b0 <- coef(lrm)[[1]] ## \beta_{0} (i.e., the model intercept) ##

x <- summary(lrm)

#######
## Writing a function for restructuring output from 
    ## "summary.glm()" as a dataframe ##
######
R.glmdf <- function(x){ ## "x" must be an object of class "summary.glm"...
                            ## ... currently only works for simple logistic
                            ## regression (i.e., only one predictor) ##
    se.cmpn <- x[c("dispersion", "cov.unscaled")]
        ## ^- "cmpn" = "components" abbreviated ##
        ## see "`r names(summary(<SOME GLM OBJECT>))`" ##
    se.a <- se.cmpn[[1]]
    se.b <- se.cmpn[[2]]
    se.ab <- se.a*se.b ## "covmat" ##
    se.c <- diag(se.ab) ## "var.cf" ##
    se.x <- sqrt(se.c) ## "s.err" ##
    z.a <- c(x$coefficients[[1]], x$coefficients[[2]])
    z.x <- z.a / se.x
    p.x <- 2 * pnorm(-abs(z.x))
    x.df <- data.frame(Estimate = c(x$coefficients[[1]],
                                    x$coefficients[[2]]),
                       SE = se.x,
                       Z = z.x,
                       p.value = p.x)
    return(x.df)
}


x.summ <- R.glmdf(x)
```

```{r lrmFits}
## **Null Deviance** ##
nulldev <- lrm$null.deviance
nulldev.df <- lrm$df.null
nulldev.p <- c(round(nulldev, 2), nulldev.df)

## **Residual Deviance** ##
dev <- lrm$deviance
dev.df <- lrm$df.residual
dev.p <- c(round(dev, 2), dev.df)

## **AIC** ##
aic <- lrm$aic
aic.p <- c(round(aic, 2), " ")

fits <- as.data.frame(rbind(nulldev.p, dev.p, aic.p))
rownames(fits) <- c("**Null Deviance**", "**Residual Deviance**", "**AIC**")
names(fits) <- c("Estimate", "Degrees of Freedom")

```

```{r lrmORCI}

CI.b <- confint(lrm, trace = FALSE)

OR <- exp(coef(lrm)) 
CI <- confint(lrm)

CIb <- cbind(coef(lrm), CI.b)
CIb.labs <- c("**$\\beta$**", dimnames(confint(lrm))[[2]])
CIb <- rbind(CIb.labs, round(CIb, 4))
dimnames(CIb)[[1]][1] <- " "
dimnames(CIb)[[2]] <- c(" ", "$CI_{\\beta}$", " ")

## odds ratios and 95% CI ##
ORCI <- exp(cbind(coef(lrm), CI))
ORCI.labs <- c("**$\\Phi$**", dimnames(confint(lrm))[[2]])
ORCI <- rbind(ORCI.labs, round(ORCI, 4))
dimnames(ORCI)[[1]][1] <- " "
dimnames(ORCI)[[2]] <- c(" ", "$CI_{\\Phi}$", " ")


library(lmtest) ## "lrtest()" ##
lrchsq <- lrtest(lrm)[2, -3]
lrchsq <- lrchsq[, c(2, 1, 3, 4)]
names(lrchsq) <- c("Log Likelihood", "_df_", "$\\chi^{2}$", "_p_")

library(modEvA)
rsq <- RsqGLM(model = lrm) %>% sapply(round, 4)
pander(rsq)
rsq.perc <- (rsq[2]*100)
```



`r tufte::newthought("Logistic Regression Analysis Summary.")` A logistic regression analysis was conducted to test whether pedestrian race^[_African American_ or _White_] predicted whether individuals were stopped by Portland Police for _major_, versus _minor_, offenses. Results for the model indicated that the severity of offenses was not predicted according to the two levels of pedestrians' race, likelihood ratio $\chi^{2}(1) = `r lrchsq[1, 3]`$ (n.s.)<!--, $\\textit{b} = `r round(lrm.b1, 3)`$ (n.s.)-->. The _Nagelkerke pseudo_-$R^{2}$ indicated that $< 1\%$ of the variance in offense severity was accounted for by the race of pedestrians stopped by PPD^[_pseudo_-$R_{2} = `r paste0(rsq[[2]]*100, "\\%")`$]. The model estimates and fit indices are summarized below.

-----

`r tufte::newthought("Logistic Regression Model Summary Statistics and Fit Indices")`

```{r lrmResults1, results='asis'}
kable(x.summ, caption = "Logistic Regression Model Summary")
# chsq
kable(lrchsq, row.names = FALSE, caption = "Likelihood Ratio $\\chi^{2}$")
```

`r tufte::newthought("Confidence Intervals (CI) \\& Odds Ratios (OR)")`

```{r lrmresults2, results='asis'}
kable(CIb, align = rep('r', ncol(CIb)), 
      caption = "Logistic Regression Coefficients ($\\beta$) \\& 
      Coresponding Confidence Intervals (_CI_)")

kable(ORCI, align = rep('r', ncol(ORCI)), 
      caption = "Logistic Regression Odds Ratios ($\\Phi$) \\& 
      Coresponding Confidence Intervals (_CI_) [note]") %>%
    add_footnote("Confidence intervals are based on the logistic regression 
                 model's profiled log-likelihood function, 
                 rather than the standard errors", 
                 threeparttable = TRUE)
```

`r tufte::newthought("Additional Model Fit Indices")`

```{r lrmresults3, results='asis'}
kable(fits, caption = "Logistic Regression Model Fit Statistics")
```


\newpage
# $1.c.$ Previous Analysis \& Results^[\todo]

`r tufte::newthought("Look back at your findings from the analysis of the contingency table in")` \textbf{\color{dkpatriarch}{\texttt{HW-1 Problem-3c}}}. Briefly compare your results from HW-1 Problem-3c to the results from the loglinear analysis and the logistic regression model^[_no more than 2-3 sentences is needed_]. Refer to specific values when making your comparisons.

`r tufte::newthought("Contingency Table \\& Group Comparisons via Pearson $\\chi^{2}$")`^[Original analysis conducted for _**Homework-1, Question-3c**_]


```{r dat2, fig.asp=1}
dat2 <- read.spss("data/race2.sav", to.data.frame = TRUE)
dat2 <- droplevels(na.omit(dat2))

tbl <- ftable(dat2$ped, dat2$offense)
tbl.df <- as.data.frame(tbl)
names(tbl.df) <- c("Race", "Offense", "Freq")

xtbl <- xtabs(Freq ~ Offense + Race, data = tbl.df, 
              sparse = TRUE, subset = NULL, na.action = NULL)
xtbl.m <- as.matrix(xtbl)
xsq <- chisq.test(xtbl.m)

library(descr) ## "CrossTable()" ##

ctab <- CrossTable(tbl.df$Race, tbl.df$Offense,
                   digits = list(expected = 3, prop = 3, percent = 3, others = 3),
                   expected = F,
                   prop.r = F, prop.c = F, prop.t = F, prop.chisq = F, 
                   chisq = T, fisher = F, mcnemar = F,
                   dnn = c("Race", "Frequency of Stops"), cell.layout = F,
                   row.labels = T,
                   total.r = T, total.c = T,
                   xlab = expression(italic("Frequency of Stops")),
                   ylab = expression(italic("Race")))

library(vcd) ## "assocstats()" ##
assocstats(tbl)
```


