---
title: "HW1 - Question 3 R Setup"
author: "Rachel M. Smith"
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

-----


# Change the file path only:

```{r echo=TRUE, results='markup'}
library(foreign)
mydata <- read.spss("data/race2.sav", to.data.frame = TRUE)
```

## The run this exactly:

```{r}
R.isna <- function(x) sum(is.na(x))
sapply(mydata, R.isna)

R.na <- function(x, v = 0){
    x <- ifelse(is.na(x), v, x)
    return(x)
}

unique(mydata$offense)
mydata <- within(mydata, {
    offense.n <- sapply(offense, R.na, v = 0)
    offense <- factor(offense.n, levels = c(0, 1, 2), labels = c("unknown", "major", "minor"), ordered = FALSE)
})
unique(mydata$offense)
sapply(mydata, R.isna)

summary(mydata)
```

-----

# Here is a breakdown of the above `R-code`:

1. Load the "`foreign`" package to read and write `spss` (among others) datafiles.
2. Load the dataset, "`race2.sav`" into `R` and save it as a `data.frame` `R`-object named "`mydata`".


```{r echo=TRUE, results='markup'}
library(foreign)
mydata <- read.spss("data/race2.sav", to.data.frame = TRUE)
```

3. Define a new function, "`R.isna()`", for getting a count of `NAs` in a given set of values (`x`).
4. Use the `sapply()` function (from the `base` package) to "simply apply" the `R.isna()` function to each column in `mydata` (i.e., this will return a table containing the sum of `NAs` in each column).


```{r echo=TRUE, results='markup'}
R.isna <- function(x) sum(is.na(x))
sapply(mydata, R.isna)
```

5. Define another new function, `R.na()`, for replacing `NAs` in a given set of values (`x`) with a specified non-`NA` value (`v`).


```{r echo=TRUE, results='markup'}
R.na <- function(x, v = 0){
    x <- ifelse(is.na(x), v, x)
    return(x)
}
```

6. Use the `unique()` function (from the `base` package) to get a list of the values currently in use for the variable being manipulated (in this case, `mydata$offense`).


```{r echo=TRUE, results='markup'}
unique(mydata$offense)
mydata <- within(mydata, {
    offense.n <- sapply(offense, R.na, v = 0)
    offense <- factor(offense.n, levels = c(0, 1, 2), labels = c("unknown", "major", "minor"), ordered = FALSE)
})
unique(mydata$offense)
sapply(mydata, R.isna)

summary(mydata)
```

-----
